Acceleration of Derivative Calculations with Application to Radial Basis Function -- Finite-Differences on the Intel MIC Architecture

### 2014-01-18 acm/ieee_paper.tex
Final paper submitted to ACM Conference. http://www.lrr.in.tum.de/~gerndt/ICS2014/

% A category with the (minimum) three required fields
\category{H.4}{Information Systems Applications}{Miscellaneous}
%A category including the fourth, optional field follows...
\category{D.2.8}{Software Engineering}{Metrics}[complexity measures, performance measures]

\terms{Theory}

@@

\ee{The core of the technique is to load the maximum amount of data
  into large register and format the data for processing within these
  registers. That way the amount of interaction with the memory is
  minimized. The Intel MIC is more for these operations than classical
  CPUs because it features larger vector registers and more elaborate
  vector reorganization instructions than namely swizzling,
  permutations and broadcasts. (The AVX instruction sets on classical
  CPUs is more primitive in that aspect.)} A vector of 512 bits is
composed of four lanes (sometimes called channels) of 128 bits, which
are made of four segments of 32 bits. A swizzling operation allows the
reorder of the elements within each lane (see how {\tt mat} is
extracted from {\tt mat\_ent} in Figure~\ref{code:mat_mul} line VIII to X). On the
other hand, a permutation reorders entire lanes, without changing
their contents (see for instance how {\tt vect} is generated from {\tt
  all\_vect} in Figure~\ref{code:mat_mul} line VII to IX).  Both permutation and
swizzling support common reordering and broadcast. In the code, the
``permute'' function needs to use the {\tt
  \_mm512\_permute4f128\_epi32} instruction which is only defined on
integers and therefore in place typecasts are necessary to convert a
vector to and from integer type.  (Note that we prefer swizzling to
permutation if possible because we believe it is cheaper. Many
instructions have swizzling capability, while permutation is its own
instruction.)

@@

\ee{The modelization of Section~\ref{sec:model} shows that the memory
  transfer caps the computation at 210GFlop/s, which is 10\% of the
  peak single precision of the architecture. To reach this value, the
  cores need to execute one "useful" Fused Multiply-Add operation
  every ten cycles. The inner loop of the kernel spends 7 vector
  instructions to layout the data and then 3 per tensor product for a
  total of 12 vector instructions. Most of these instruction have a
  one cycle throughput cost except the gather operations which might
  need to perform up to 16 memory accesses. There are also some scalar
  operations but there are few of them and are executed on a different
  pipeline than the vector operations. Overall, the inner loop can
  emit the 4 Fused Multiply-Add in less than 40 cycles.}

@@

  \caption{Performance of $y=Ax$ for a RBF derivative stencil of a
    $96^3$ 3D grid. Classical "one-matrix one-vector" SpMV ($n_m=n_v=1$ given in blue) has low
    performance. The manually vectorized code (red) is 3.5 times
    faster than the compiler generated one (green). RCM ordering
    (squares) only improves the performance for our manually
    vectorized code.
    We also show peak theoretical performance for the base case and the 4 vector/1 matrix case. Not shown is the peak theoretical performance of 210 Gflops for the 4/4 case. 
    }
  \label{fig:comparison_spmv}
  \label{fig:perf_mic}
\end{figure}

@@

In future work, we will examine the effect of larger stencil sizes and
double precision. The presented performance does not reach the
predicted peak and we believe that common techniques from SpMV such as
cache partitioning can be successfully applied. We will also integrate
the techniques presented in this paper within a fluid simulation using
RBFFD. \ee{We will also look developping a similar technique on more
commonly available devices such as GPUs.}

>>>>

% A category with the (minimum) three required fields
%\category{I.6}{Simulation and Modeling}{Applications} %\break
\category{F.2}{Analysis of Algorithms and Problem Complexity}{Numerical Algorithms and Problems}
%A category including the fourth, optional field follows...
%\category{D.2.8}{Software Engineering}{Metrics}[complexity measures, performance measures]

%\terms{Theory}

@@

\ee{The core of the technique is to load the maximum amount of data
  into large the registers and format the data for processing within these
  registers, minimizing the amount of interaction with memory. 
  The Intel MIC is more for these operations than classical
  CPUs because it features larger vector registers and more elaborate
  vector reorganization instructions, such as swizzling,
  permutations and broadcasts. (The AVX instruction sets on classical
  CPUs are more primitive in that aspect.)} A vector of 512 bits is
composed of four lanes (sometimes called channels) of 128 bits, which
are made of four segments of 32 bits. A swizzling operation allows the
reorder of the elements within each lane (see how {\tt mat} is
extracted from {\tt mat\_ent} in Figure~\ref{code:mat_mul} line VIII to X). On the
other hand, a permutation reorders entire lanes, without changing
their contents (see for instance how {\tt vect} is generated from {\tt
  all\_vect} in Figure~\ref{code:mat_mul} line VII to IX).  Both permutation and
swizzling support common reordering and broadcast. In the code, the
``permute'' function needs to use the {\tt
  \_mm512\_permute4f128\_epi32} instruction which is only defined on
integers and therefore in place typecasts are necessary to convert a
vector to and from integer type.  (Note that we prefer swizzling to
permutation if possible because we believe it is cheaper. Many
instructions have swizzling capability, while permutation is its own
instruction.)

@@

\ee{The modelization of Section~\ref{sec:model} establishes that the 
  memory
  transfer caps the computation at 210~GFlop/s, which is 10\% of the
  peak single precision of the architecture. To reach this value, the
  cores need to execute one "useful" Fused Multiply-Add operation
  every ten cycles. The inner loop of the kernel spends 7 vector
  instructions to layout the data and then 3 per tensor product for a
  total of 12 vector instructions. Most of these instruction have a
  one cycle throughput cost except the gather operations which might
  need to perform up to 16 memory accesses. There are also some scalar
  operations but there are few of them and are executed on a different
  pipeline than the vector operations. Overall, the inner loop can
  emit the 4 Fused Multiply-Add in less than 40 cycles.}

@@

  \caption{Performance of $y=Ax$ for a RBF derivative stencil of a
    $96^3$ 3D grid. Classical "one-matrix one-vector" SpMV ($n_m=n_v=1$ given in blue) has low
    performance. The manually vectorized code (red) is 3.5 times
    faster than the compiler generated one (green). RCM ordering
    (squares) only improves the performance for our manually
    vectorized code.
    We also show peak theoretical performance for v1m1 and v4m1 on the MIC. 
    Not shown is the peak theoretical performance of 210 Gflops for v4m4. 
    }
  \label{fig:comparison_spmv}
  \label{fig:perf_mic}
\end{figure}

@@

In future work, we will examine the effect of larger stencil sizes and
double precision. The presented performance does not reach the
predicted peak and we believe that common techniques from SpMV such as
cache partitioning can be successfully applied. We will also integrate
the techniques presented in this paper within a fluid simulation using
RBFFD. \ee{We will also look into developing a similar technique on more
commonly available devices such as GPUs.}
